Estimating duration in the presence of misclassification

### 2024-03-22 cis-imperfect-testing.tex

Comparing repeated data simulations, using the set-up of \cref{E-perf-test:sec:simulation-study}, to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
Specifically, if $j$ is a single positive episode then $j$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
% This produced the same pattern, that is estimating too many short episodes (not shown).

@@

The CIS is a more complex setting, including double interval censoring and a complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
This means the result cannot simply be extended to the CIS, but still reinforces the intuition that single positive episodes are likely to be problematic.

@@

First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being missed.

@@

Denote the replacement for $p_{iu}$ as $p_{iu}'$.
$p_{iu}'$ is the probability of episode $i$ being missed, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
These mechanisms are mutually exclusive.
Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \eta$.
For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1) \right) \\
&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}

@@

Next, I considered the consequence of $\psens$ being misspecified, I use the model combination prior because it is the better performing prior in the correctly specified case.
If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates.
The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is overestimated.
This effect dominates over the opposing bias of overestimating the number of missed episodes.
The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
\begin{figure}
  % \makebox[\textwidth][c]{
    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-misspecified-sensitivity}
  \caption[Simulation study results with misspecified test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
    The true survival time is shown in black
    All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
    Hence, (B) is correctly specified but (A) and (C) have misspecified $\psens$.
  }
  \label{imperf-test:fig:misspecified-test-sensitivity}
\end{figure}

@@

An informative prior on $\ntot$ is required is for the bulk of the distribution to agree with those in \cref{E-ATACCC} and the wider literature.
In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
A possible explanation is that the true changes in the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
In particular, the test sensitivity for long episodes may become much lower than 50\%.
This allows many more episodes to be missed than the model assumes, and hence a higher $\ntot$ is required to explain the data.
Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to when the simulating using a too low test sensitivity.
Violating these assumptions would lead to unpredictable inference results, perhaps those seen here.

@@

% However, comparing the CIS and simulated datasets do not show any differences which suggest what in the data is causing these differences.
% Then, the additional missed infections implied by a 
% This would lead to more missed infections than the model assumes, and hence a higher $\ntot$ is required to explain the data.

>>>>

Comparing repeated data simulations, using the set-up of \cref{E-perf-test:sec:simulation-study}, to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
Specifically, if $j$ is a single positive episode then $j$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
Additionally, short episodes are very likely to be undetected, which means that the design of the CIS cannot rule out many of them occurring.
% This produced the same pattern, that is estimating too many short episodes (not shown).

@@

The CIS is a more complex setting, including double interval censoring and a complex pattern of undetected episodes (see \cref{E-perf-test:sec:problem}).
This means the result cannot simply be extended to the CIS, but still reinforces the intuition that single positive episodes are likely to be problematic.

@@

First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being undetected.

@@

Denote the replacement for $p_{iu}$ as $p_{iu}'$.
$p_{iu}'$ is the probability of episode $i$ being undetected, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
These mechanisms are mutually exclusive.
Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \eta$.
For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
\begin{align}
1 - p_{iu}'
&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1) \right) \\
&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1)\right).
\label{imperf-test:eq:pit-prime}
\end{align}

@@

Next, I considered the consequence of $\psens$ being misspecified, I use the model combination prior because it is the better performing prior in the correctly specified case.
If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates.
The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is overestimated.
This effect dominates over the opposing bias of overestimating the number of undetected episodes.
The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
\begin{figure}
  % \makebox[\textwidth][c]{
    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-misspecified-sensitivity}
  \caption[Simulation study results with misspecified test sensitivity]{%
    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
    The true survival time is shown in black
    All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
    Hence, (B) is correctly specified but (A) and (C) have misspecified $\psens$.
  }
  \label{imperf-test:fig:misspecified-test-sensitivity}
\end{figure}

@@

An informative prior on $\ntot$ is required is for the bulk of the distribution to agree with those in \cref{E-ATACCC} and the wider literature.
In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
A possible explanation is that the true changes in the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
In particular, the test sensitivity for long episodes may become much lower than 50\%.
This allows many more episodes to be undetected than the model assumes, and hence a higher $\ntot$ is required to explain the data.
Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to when the simulating using a too low test sensitivity.
Violating these assumptions would lead to unpredictable inference results, perhaps those seen here.

@@

% However, comparing the CIS and simulated datasets do not show any differences which suggest what in the data is causing these differences.
% Then, the additional undetected infections implied by a 
% This would lead to more undetected infections than the model assumes, and hence a higher $\ntot$ is required to explain the data.

### 2024-03-22 introduction.tex
minor to dos

Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be missed (a generalisation of truncation).
This is based on the CIS data.
While statistical frameworks have been previously developed which included this case, they had not been used in practice.
In \cref{E-imperf-test}, I extend this framework to include false negatives.

@@

In order to best fulfil this thesis's primary aim, the methodology developed should be easy for others to use.
Therefore, the code for the analyses in this thesis is available online.
An index to the various repositories is available at\todo{insert GitHub link}.
I structured two elements of the code as R packages to enable their application to future contexts.

>>>>

Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be undetected (a generalisation of truncation).
This is based on the CIS data.
While statistical frameworks have been previously developed which included this case, they had not been used in practice.
In \cref{E-imperf-test}, I extend this framework to include false negatives.

@@

In order to best fulfil this thesis's primary aim, the methodology developed should be easy for others to use.
Therefore, the code for the analyses in this thesis is available online.
An index to the code and, where possible, data is available at \url{https://github.com/joshuablake/thesis-index}.
I structured two elements of the code as R packages to enable their application to future contexts.

### 2024-03-22 transmission-appendix-INLA.tex

INLA (Integrated Nested Laplace Approximation) is a method for approximate Bayesian inference.
It was introduced by \textcite{rueINLA}.
This appendix is based on that paper, as well as the seminar \textcite{rueINLAseminar} and the tutorial paper \textcite{martinoINLAtutorial}.

@@

INLA requires a LGM (latent Gaussian model).
An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $\pi(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j \pi(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
% The Gaussian structure allows the use of the approximations in \cref{transmission:sec:INLA:inference}.
% The conditional independence structure means that the precision matrix, $\matr{Q} = \matr{\Sigma}^{-1}$, is sparse.

For INLA to perform well, \ie be compuutationally efficient and an accurate approximation, $\vec{\theta}$ should be of low dimension, \textcite{martinoINLAtutorial} recommend $< 15$; and the precision matrix, $\matr{Q}$, should be sparse.
The low dimensionality of $\vec{\theta}$ is because the method requires numerical integration over it.
The sparsity of $\matr{Q}$ allows for efficient computation.
Sparsity is introduced by conditional independence structures of the latent field that are present in many models.
For example, in the classic random effect setting, all random effects are independent conditional on the hyperparameters.

@@

Finally, the marginal posterior density of $x_i$ is found by integrating over $\vec{\theta}$:
\begin{align}
\pi(x_i \mid \vec{y})
&\approx \int \pi(x_i \mid \vec{\theta}, \vec{y}) \tilde\pi(\vec{\theta} \mid \vec{y}) d\vec{\theta}.
\end{align}
which can be computed using numerical integration over $\tilde\pi(\vec{\theta} \mid \vec{y})$, previously calculated.

@@

INLA assumes that $\vec{x} \mid \vec{y}, \vec{\theta}$ is approximately Gaussian, and therefore takes a Laplace approximation to the distribution.
The greater the smoothing, the better the approximation.
This is because, if $\vec{x} \mid \vec{y}, \vec{\theta}$ is then dominated by the prior $\vec{x} \mid \vec{\theta}$ and the prior is Gaussian.
Exploiting the structure of the problem, in particular the conditional independence structure, the Laplace approximation can be computed efficiently.

>>>>

INLA (Integrated Nested Laplace Approximation) is a method for approximate Bayesian inference for LGMs (latent Gaussian models).
LGMs are a broad class of models, including all generalised linear mixed models and generalised linear additive mixed models, as well as many more complex models.

It was introduced by \textcite{rueINLA}.
This appendix is based on that paper, as well as the seminar \textcite{rueINLAseminar} and the tutorial paper \textcite{martinoINLAtutorial}.

@@

An LGM is a model for a series of observations $\vec{y} = [y_1, \dots, y_n]^T$.
Each observation has an associated linear predictor $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
The linear predictor and, optionally, some hyperparameters $\vec{\theta}$ give the conditional independence structure:
\begin{align}
    \pi(\vec{y} \mid \vec{\eta}, \vec{\theta}) = \prod_{i=1}^n \pi(y_i \mid \eta_i, \vec{\theta}).
\end{align}
Such a structure is familiar from linear regression models which have $\vec{\eta} = \matr{X} \vec{\beta}$, where $\matr{X}$ is the design matrix and $\vec{\beta}$ are the coefficients; $\vec{\theta}$ is one-dimensional; and $\pi(y_i \mid \eta_i, \vec{\theta})$ is the normal distribution with mean $\eta_i$ and standard deviation $\theta$.

An LGM introduces a latent field $\vec{x}$ with a multivariate Gaussian distribution (hence the name):
\begin{align}
    \vec{x} \mid \vec{\theta} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\theta}))
\end{align}
where $\Sigma(\vec{\theta})$ is the covariance matrix formed as a deterministic transformation of the hyperparameters.
The linear predictors must be able to be expressed as a linear transformation of the latent field, \ie $\vec{\eta} = \matr{A} \vec{x}$ for some pre-specified matrix $\matr{A}$.
For the linear regression model, $\matr{A} = \matr{X}$, $\vec{x} = \vec{\beta}$, and $\matr{\Sigma}(\vec{\theta})$ would be a fixed diagonal matrix specifying the prior on the regression coefficients.
In the (generalised) linear model case $\vec{\theta}$ is fixed; however, LGMs allow an arbitrary prior distribution on $\vec{\theta}$ which means that some elements of $\vec{x}$ can be random effects (or more complex structures).

In summary, the full specification of an LGM is:
\begin{align}
    \vec{\theta} &\dist \pi(\vec{\theta}) \\
    \vec{x} \mid \vec{\theta} &\dist \N(\vec{0}, \matr{\Sigma}(\vec{\theta})) \\
    \vec{\eta} &= \matr{A} \vec{x} \\
    \pi(\vec{y} \mid \vec{\eta}, \vec{\theta}) &= \prod_{i=1}^n \pi(y_i \mid \eta_i, \vec{\theta})
\end{align}
where $\pi(\vec{\theta})$ is an arbitrary prior distribution on the hyperparameters.

% An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
% To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $\pi(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j \pi(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
% The Gaussian structure allows the use of the approximations in \cref{transmission:sec:INLA:inference}.
% The conditional independence structure means that the precision matrix, $\matr{Q} = \matr{\Sigma}^{-1}$, is sparse.

For INLA to perform well, \ie be compuutationally efficient and an accurate approximation, $\vec{\theta}$ should be of low dimension (\textcite{martinoINLAtutorial} recommend $< 15$); and the precision matrix, $\matr{\Sigma}^{-1}$, should be sparse.
However, the dimension of $\vec{x}$ is unrestricted, allowing a large number of random effects, or more complex structures.
The low dimensionality of $\vec{\theta}$ is because the method requires numerical integration over it.
The sparsity of $\matr{\Sigma}^{-1}$ allows for efficient computation.
Sparsity is introduced by conditional independence structures of the latent field that are present in many models.
For example, in the classic random effect setting, all random effects are independent conditional on the hyperparameters.

% \subsection{Application}

% Next, I explain why the inference-inferred approach proposed in \cref{backcalc:sec:approach} cannot be expressed as an LGM.
% In this context, each $i$ refers to a specific stratum and day combination.
% The number of positive test results in this stratum and day is $y_i$.
% $\eta_i$ is the population prevalence in this stratum on this day.
% Hence, $y_i \mid \eta_i, \rho \dist \BB(n_i, \eta_i, \rho)$ where $\rho$ is a hyperparameter (in $\vec{\theta}$) determining the overdispersion of the beta-binomial distribution and $n_i$ is the number of .

% In this context, each $w_j$ corresponds to the observed RT-PCR results in one stratum on one day.
% Hence, $\eta_{j}$ must determine the distribution of $w_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.

% In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
% If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is an LGM.
% However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
% In particular, if $\eta_j$ is the prevalence, then:
% \begin{align}
%     \eta_{i,t}
%     &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
%     &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
% \end{align}
% As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of an LGM.

@@

Finally, the marginal posterior density of $x_i$ is found by integrating over $\vec{\theta}$:
\begin{align}
\pi(x_i \mid \vec{y})
&\approx \int \pi(x_i \mid \vec{\theta}, \vec{y}) \tilde\pi(\vec{\theta} \mid \vec{y}) d\vec{\theta}
\end{align}
computed using numerical integration.

@@

INLA assumes that $\vec{x} \mid \vec{y}, \vec{\theta}$ is approximately Gaussian, and therefore takes a Laplace approximation to the distribution.
The approximation then proceeds as before.
Exploiting the structure of the problem, in particular the conditional independence structure, the Laplace approximation can be computed efficiently.

The greater the smoothing, the better the Laplace approximation.
This is because, if $\vec{x} \mid \vec{y}, \vec{\theta}$ is then dominated by the prior $\vec{x} \mid \vec{\theta}$ and the prior is Gaussian.
